---
title: "Chronic Disease Prevalence in the United States"
subtitle: "Analysis using PLACES: County Data, 2023 release"
author: "Augustine Takyi and Chetan Elenki"
output: 
    html_document:
        theme: paper
        highlight: breeze
        code_folding: show
        fig_caption: true
        toc: true
        toc_depth: 3
        toc_float: true
        number_sections: true
        fig_width: 8
        fig_height: 6
        fig_align: "center"
        df_print: paged
        keep_md: true
---

# Introduction
The United States has a high prevalence of chronic diseases such as heart disease, cancer, and diabetes. 
These diseases are the leading causes of death and disability in the United States. 
The Centers for Disease Control and Prevention (CDC) estimates that chronic diseases account for 7 out of 10 deaths in the United States. 
The CDC also estimates that chronic diseases are responsible for 86% of the nation's healthcare costs. 
The prevalence of chronic diseases in the United States is a major public health concern.

The CDC's PLACES project provides county-level data on chronic disease prevalence, health outcomes, and health behaviors in the United States.
The PLACES project aims to provide data that can be used to identify and address health disparities at the local level.
The PLACES data can be used to identify areas with high rates of chronic diseases and to develop targeted interventions to improve health outcomes.

In this analysis, we will use the PLACES: County Data, 2023 release to examine the prevalence of chronic diseases in the United States.
We will analyze the prevalence of chronic diseases such as arthritis, high blood pressure, cancer, asthma, and diabetes.
We will also examine other health outcomes, health behaviors, and social determinants of health in the United States.
The goal of this analysis is to identify areas with high rates of chronic diseases and to explore the factors that contribute to these disparities.

# Data Source Description
For this analysis, we will use PLACES: Local Data for Better Health, Place Data 2023 release.
Here's a brief description of the dataset (as provided by the CDC, source cited below):

* This dataset contains model-based county estimates. 
* PLACES covers the entire United States—50 states and the District of Columbia—at county, place, census tract, and ZIP Code Tabulation Area levels. 
* It provides information uniformly on this large scale for local areas at four geographic levels. 
* Estimates were provided by the Centers for Disease Control and Prevention (CDC), Division of Population Health, Epidemiology and Surveillance Branch. 
* PLACES was funded by the Robert Wood Johnson Foundation in conjunction with the CDC Foundation. 

* This dataset includes estimates for 36 measures: 
    * 13 for health outcomes, 
    * 10 for preventive services use, 
    * 4 for chronic disease-related health risk behaviors, 
    * 7 for disabilities, and 3 for health status. 

* Data sources used to generate these model-based estimates are:
    * Behavioral Risk Factor Surveillance System (BRFSS) 2021 or 2020 data, 
    * Census Bureau 2021 or 2020 county population estimate data, and 
    * American Community Survey 2017–2021, or 2016–2020 estimates. 

* The 2023 release uses 2021 BRFSS data for 29 measures and 2020 BRFSS data for 7 measures 
    (all teeth lost, dental visits, mammograms, cervical cancer screening, colorectal cancer screening, core preventive services among older adults, and sleeping less than 7 hours) 
    that the survey collects data on every other year. 

* More information about the methodology can be found at www.cdc.gov/places.

**Source:** 
    
* https://data.cdc.gov/500-Cities-Places/PLACES-Local-Data-for-Better-Health-Place-Data-202/eav7-hnsx/about_data

* Further information about the dataset can be found here: https://www.cdc.gov/places/about/index.html

# Exploration and Data Preparation
## Quick glimpse into the data
Lets start by loading the data and examining the first few rows.
```{r load_data, message=FALSE, warning=FALSE}
    # loading data
    library(readr)

    # google drive location of the files
    places_county_2023_url <- 'https://drive.google.com/file/d/1XN8AfQ36pSBkfFXMI3rLpjij0Rt67WCj/view?usp=sharing'

    # define a function to read the data
    read_data <- function(url){
    # extract file id from the url
    file_id <- strsplit(url, '/')[[1]][6]
    
    # create a read path
    read_path <- paste0('https://drive.usercontent.google.com/download?id=', file_id, '&export=download&authuser=0&confirm=t')
    
    # read the data
    df <- read_csv(read_path)
    return(df)
    }

    # read the data
    places_county_2023 <- read_data(places_county_2023_url)

    # display the first few rows
    head(places_county_2023)
```

The dataset contains the following columns:
```{r column_names, message=FALSE, warning=FALSE}
    # display the column names
    colnames(places_county_2023)
```

## Features in the dataset
The dataset contains a wide range of features related to chronic diseases, health outcomes, health behaviors, and social determinants of health.
Here are some of the key features in the dataset:

```{r load_libraries, message=FALSE, warning=FALSE}
    library(dplyr)
    library(tidyr)
```

### US Counties Data
The dataset is at the county level and includes information about the state, county name, county FIPS code, total population, and geolocation.
All observations in the dataset are identified by the following key columns:
```{r key_columns, message=FALSE, warning=FALSE}
    key_cols <- c('StateAbbr', 'StateDesc', 'CountyName', 'CountyFIPS', 'TotalPopulation', 'Geolocation')
    places_county_2023[key_cols] %>% head()
```

We will start by preparing counties dataset, as this is the granularity level for all our data.
```{r counties_data_preparation, message=FALSE, warning=FALSE}
    # using key_cols to filter the required columns
    counties_data <- places_county_2023[key_cols] %>% distinct()

    # check for missing values
    missing_values <- sapply(counties_data, function(x) sum(is.na(x)))
    missing_values
```
As we can see, no missing values are present in the counties dataset.

For our analysis, we will enrich the counties dataset with additional information about the location type - whether it is a city, town, or rural area.
This will be useful for understanding the context of the counties and their health outcomes.
```{r enrich_county_data, message=FALSE, warning=FALSE}
    # load urban rural classification data
    urban_rural_url <- 'https://drive.google.com/file/d/1-w2AvZj4_sqEzMMS-NCPsILAZzKPaVvE/view?usp=sharing'
    urban_rural_data <- read_data(urban_rural_url)

    # select the required columns
    # 'State Abr.', 'County name', 'FIPS code', '2013 code'
    # rename to: 'StateAbbr', 'LocationName', 'FIPS', 'urban_rural_code'
    # stip 'County' from 'County name'
    urban_rural_data <- urban_rural_data %>%
        select(`State Abr.` = 'State Abr.', `County name` = 'County name', `FIPS code` = 'FIPS code', `2013 code` = '2013 code') %>%
        rename(StateAbbr = `State Abr.`, LocationName = `County name`, FIPS = `FIPS code`, urban_rural_code = `2013 code`) %>%
        mutate(LocationName = gsub(' County', '', LocationName))

    # transform FIPS column
    # if length of FIPS is less than 5, add leading zeros
    urban_rural_data$FIPS <- sprintf('%05d', urban_rural_data$FIPS)

    # create a new column 'urban_rural_classification' based on 'urban_rural_code'
    # 1 - "Large central metro", 2 - "Large fringe metro", 3 - "Medium metro", 4 - "Small metro", 5 - "Micropolitan", 6 - "Noncore"
    urban_rural_data$urban_rural_classification <- case_when(
        urban_rural_data$urban_rural_code == 1 ~ 'Large central metro',
        urban_rural_data$urban_rural_code == 2 ~ 'Large fringe metro',
        urban_rural_data$urban_rural_code == 3 ~ 'Medium metro',
        urban_rural_data$urban_rural_code == 4 ~ 'Small metro',
        urban_rural_data$urban_rural_code == 5 ~ 'Micropolitan',
        urban_rural_data$urban_rural_code == 6 ~ 'Noncore'
    )

    head(urban_rural_data)

    # join urban_rural_data with key_cols and get the urban_rural_classification
    # join on 'FIPS' in urban_rural_data and 'CountyFIPS' in key_cols
    enriched_county_data <- counties_data %>%
        left_join(urban_rural_data %>% select(FIPS, urban_rural_code, urban_rural_classification), by = c('CountyFIPS' = 'FIPS'))

    # display the county data
    head(enriched_county_data)

    # group by urban_rural_classification and get the count
    urban_rural_count <- enriched_county_data %>%
        group_by(urban_rural_classification) %>%
        summarise(count = n())

    # display the urban_rural_count
    urban_rural_count

    # display where urban_rural_classification is missing
    missing_urban_rural <- enriched_county_data %>%
        filter(is.na(urban_rural_classification))
    missing_urban_rural

    # the 2 counties we are missing urban_rural_classification for are: 'Copper River' and 'Chugach' counties both in Alaska
    # we will manually set the urban_rural_code and urban_rural_classification for these counties
    enriched_county_data <- enriched_county_data %>%
        mutate(
            urban_rural_code = case_when(
                CountyName == 'Copper River' ~ 6,
                CountyName == 'Chugach' ~ 6,
                TRUE ~ urban_rural_code
            ),
            urban_rural_classification = case_when(
                CountyName == 'Copper River' ~ 'Noncore',
                CountyName == 'Chugach' ~ 'Noncore',
                TRUE ~ urban_rural_classification
            )
        )

    # check for missing values
    missing_values <- sapply(enriched_county_data, function(x) sum(is.na(x)))
    missing_values

    # finally, for ease of joining with other datasets, we will create a LocationID column
    # which is a combination of StateAbbr and CountyFIPS
    enriched_county_data$LocationID <- paste(enriched_county_data$StateAbbr, enriched_county_data$CountyFIPS, sep = '_')

    # urban_rural_code and urban_rural_classification are now available for all counties
    # we can convert urban_rural_code to a factor for better analysis
    enriched_county_data$urban_rural_code <- as.factor(enriched_county_data$urban_rural_code)
    enriched_county_data$urban_rural_classification <- as.factor(enriched_county_data$urban_rural_classification)

    # display the enriched county data
    head(enriched_county_data)
```

### Chronic Disease Data
The dataset includes several features related to chronic diseases such as arthritis, high blood pressure, cancer, asthma, and diabetes.
Here are some of the key features related to chronic diseases:
```{r chronic_disease_columns, message=FALSE, warning=FALSE}
    chronic_disease_cols <- c('ARTHRITIS_AdjPrev', 'BPHIGH_AdjPrev', 'CANCER_AdjPrev', 'CASTHMA_AdjPrev', 
                                'CERVICAL_AdjPrev', 'CHD_AdjPrev', 'COPD_AdjPrev', 'DEPRESSION_AdjPrev', 
                                'DIABETES_AdjPrev', 'KIDNEY_AdjPrev', 'STROKE_AdjPrev')
    places_county_2023[chronic_disease_cols] %>% head()
```

Next, we will prepare the chronic disease data by filtering the required columns and handling missing values.
```{r chronic_disease_data_preparation, message=FALSE, warning=FALSE}
    # for this dataset we will select the key_cols and chronic_disease_cols
    chronic_disease_data <- places_county_2023[key_cols %>% union(chronic_disease_cols)] %>% distinct()

    # create a LocationID column and drop key_cols
    chronic_disease_data$LocationID <- paste(chronic_disease_data$StateAbbr, chronic_disease_data$CountyFIPS, sep = '_')
    chronic_disease_data <- chronic_disease_data %>% select(-key_cols)

    # check for missing values
    missing_values <- sapply(chronic_disease_data, function(x) sum(is.na(x)))
    missing_values

    # as the number of missing values is small, we will drop the rows with missing values
    chronic_disease_data <- chronic_disease_data %>% drop_na()

    # display the chronic disease data
    head(chronic_disease_data)
```

### Health Status Data
The dataset also includes features related to general health status, mental health, and physical health.
Here are some of the key features related to general health:
```{r general_health_columns, message=FALSE, warning=FALSE}
    health_status_cols <- c('GHLTH_AdjPrev', 'MHLTH_AdjPrev', 'PHLTH_AdjPrev')
    places_county_2023[health_status_cols] %>% head()
```

Next, we will prepare the health status data by filtering the required columns and handling missing values.
```{r health_status_data_preparation, message=FALSE, warning=FALSE}
    # for this dataset we will select the key_cols and health_status_cols
    health_status_data <- places_county_2023[key_cols %>% union(health_status_cols)] %>% distinct()

    # create a LocationID column and drop key_cols
    health_status_data$LocationID <- paste(health_status_data$StateAbbr, health_status_data$CountyFIPS, sep = '_')
    health_status_data <- health_status_data %>% select(-key_cols)

    # check for missing values
    missing_values <- sapply(health_status_data, function(x) sum(is.na(x)))
    missing_values

    # as the number of missing values is small, we will drop the rows with missing values
    health_status_data <- health_status_data %>% drop_na()

    # display the health status data
    head(health_status_data)
```

### Risk Factors Data
The dataset includes features related to risk factors for chronic diseases such as high cholesterol and obesity.
Here are some of the key features related to risk factors:
```{r risk_factors_columns, message=FALSE, warning=FALSE}
    risk_factors_cols <- c('HIGHCHOL_AdjPrev', 'OBESITY_AdjPrev')
    places_county_2023[risk_factors_cols] %>% head()
```

Next, we will prepare the risk factors data by filtering the required columns and handling missing values.
```{r risk_factors_data_preparation, message=FALSE, warning=FALSE}
    # for this dataset we will select the key_cols and risk_factors_cols
    risk_factors_data <- places_county_2023[key_cols %>% union(risk_factors_cols)] %>% distinct()

    # create a LocationID column and drop key_cols
    risk_factors_data$LocationID <- paste(risk_factors_data$StateAbbr, risk_factors_data$CountyFIPS, sep = '_')
    risk_factors_data <- risk_factors_data %>% select(-key_cols)

    # check for missing values
    missing_values <- sapply(risk_factors_data, function(x) sum(is.na(x)))
    missing_values

    # as the number of missing values is small, we will drop the rows with missing values
    risk_factors_data <- risk_factors_data %>% drop_na()

    # display the risk factors data
    head(risk_factors_data)
```

### Disabilities Data
The dataset includes features related to disabilities such as disability status and disability prevalence.
Here are some of the key features related to disabilities:
```{r disability_columns, message=FALSE, warning=FALSE}
    disabilities_cols <- c('DISABILITY_AdjPrev', 'HEARING_AdjPrev', 'VISION_AdjPrev', 'COGNITION_AdjPrev', 'MOBILITY_AdjPrev', 'SELFCARE_AdjPrev', 'INDEPLIVE_AdjPrev')
    places_county_2023[disabilities_cols] %>% head()
```

Next, we will prepare the disabilities data by filtering the required columns and handling missing values.
```{r disabilities_data_preparation, message=FALSE, warning=FALSE}
    # for this dataset we will select the key_cols and disabilities_cols
    disabilities_data <- places_county_2023[key_cols %>% union(disabilities_cols)] %>% distinct()

    # create a LocationID column and drop key_cols
    disabilities_data$LocationID <- paste(disabilities_data$StateAbbr, disabilities_data$CountyFIPS, sep = '_')
    disabilities_data <- disabilities_data %>% select(-key_cols)

    # check for missing values
    missing_values <- sapply(disabilities_data, function(x) sum(is.na(x)))
    missing_values

    # as the number of missing values is small, we will drop the rows with missing values
    disabilities_data <- disabilities_data %>% drop_na()

    # display the disabilities data
    head(disabilities_data)
```

### Health Behaviors Data
The dataset includes features related to health behaviors such as smoking, physical inactivity, and excessive drinking, which are negative health behaviors that can contribute to chronic diseases.
```{r negative_health_behaviors, message=FALSE, warning=FALSE}
    negative_health_behavior_cols <- c('BINGE_AdjPrev', 'CSMOKING_AdjPrev')
    places_county_2023[negative_health_behavior_cols] %>% head()
```

The dataset also includes positive health behaviors such as physical activity.
```{r positive_health_behaviors, message=FALSE, warning=FALSE}
    positive_health_behavior_cols <- c('LPA_AdjPrev')
    places_county_2023[positive_health_behavior_cols] %>% head()
```

Next, we will prepare the health behaviors data by filtering the required columns and handling missing values.
```{r health_behaviors_data_preparation, message=FALSE, warning=FALSE}
    # negative_health_behavior_cols
    # for this dataset we will select the key_cols and negative_health_behavior_cols
    negative_health_behaviors_data <- places_county_2023[key_cols %>% union(negative_health_behavior_cols)] %>% distinct()

    # create a LocationID column and drop key_cols
    negative_health_behaviors_data$LocationID <- paste(negative_health_behaviors_data$StateAbbr, negative_health_behaviors_data$CountyFIPS, sep = '_')
    negative_health_behaviors_data <- negative_health_behaviors_data %>% select(-key_cols)

    # check for missing values
    missing_values <- sapply(negative_health_behaviors_data, function(x) sum(is.na(x)))
    missing_values

    # as the number of missing values is small, we will drop the rows with missing values
    negative_health_behaviors_data <- negative_health_behaviors_data %>% drop_na()

    # display the negative health behaviors data
    head(negative_health_behaviors_data)

    # positive_health_behavior_cols
    # for this dataset we will select the key_cols and health_behaviors_cols
    positive_health_behaviors_data <- places_county_2023[key_cols %>% union(positive_health_behavior_cols)] %>% distinct()

    # create a LocationID column and drop key_cols
    positive_health_behaviors_data$LocationID <- paste(positive_health_behaviors_data$StateAbbr, positive_health_behaviors_data$CountyFIPS, sep = '_')
    positive_health_behaviors_data <- positive_health_behaviors_data %>% select(-key_cols)

    # check for missing values
    missing_values <- sapply(positive_health_behaviors_data, function(x) sum(is.na(x)))
    missing_values

    # as the number of missing values is small, we will drop the rows with missing values
    positive_health_behaviors_data <- positive_health_behaviors_data %>% drop_na()

    # display the positive health behaviors data
    head(positive_health_behaviors_data)
```

### Preventive Care Data
The dataset includes features related to preventive care such as mammography screening, regular checkups, etc.
```{r preventive_care_columns, message=FALSE, warning=FALSE}
    preventive_care_cols <- c('BPMED_AdjPrev', 'CHECKUP_AdjPrev', 'CHOLSCREEN_AdjPrev', 'COLON_SCREEN_AdjPrev')
    places_county_2023[preventive_care_cols] %>% head()
```
Next, we will prepare the preventive care data by filtering the required columns and handling missing values.
```{r preventive_care_data_preparation, message=FALSE, warning=FALSE}
    # for this dataset we will select the key_cols and preventive_care_cols
    preventive_care_data <- places_county_2023[key_cols %>% union(preventive_care_cols)] %>% distinct()

    # create a LocationID column and drop key_cols
    preventive_care_data$LocationID <- paste(preventive_care_data$StateAbbr, preventive_care_data$CountyFIPS, sep = '_')
    preventive_care_data <- preventive_care_data %>% select(-key_cols)

    # check for missing values
    missing_values <- sapply(preventive_care_data, function(x) sum(is.na(x)))
    missing_values

    # as the number of missing values is small, we will drop the rows with missing values
    preventive_care_data <- preventive_care_data %>% drop_na()

    # display the preventive care data
    head(preventive_care_data)
```

### Social Determinants Data
We have only one feature related to social determinants of health in the dataset, which is access to health insurance.
```{r social_determinants_columns, message=FALSE, warning=FALSE}
    sdoh_cols <- c('ACCESS2_AdjPrev')
    places_county_2023[sdoh_cols] %>% head()
```

Next, we will prepare the social determinants data by filtering the required columns and handling missing values.
```{r social_determinants_data_preparation, message=FALSE, warning=FALSE}
    # for this dataset we will select the key_cols and sdoh_cols
    social_determinants_data <- places_county_2023[key_cols %>% union(sdoh_cols)] %>% distinct()

    # create a LocationID column and drop key_cols
    social_determinants_data$LocationID <- paste(social_determinants_data$StateAbbr, social_determinants_data$CountyFIPS, sep = '_')
    social_determinants_data <- social_determinants_data %>% select(-key_cols)

    # check for missing values
    missing_values <- sapply(social_determinants_data, function(x) sum(is.na(x)))
    missing_values

    # as the number of missing values is small, we will drop the rows with missing values
    social_determinants_data <- social_determinants_data %>% drop_na()

    # display the social determinants data
    head(social_determinants_data)
```

# Data Analysis
For our analysis, we theorize that the age-adjusted prevalence of chronic diseases is influenced by:
    * whether the county is urban or rural, 
    * the prevalence of disabilities, 
    * negative health behaviors, 
    * positive health behaviors, 
    * preventive care, 
    * and social determinants of health. 

We will analyze the relationship between these factors and the age-adjusted prevalence of chronic diseases.

## Influence of the county type on the age-adjusted prevalence of chronic diseases
We will analyze the influence of the county type (urban or rural) on the age-adjusted prevalence of chronic diseases.
```{r county_type_analysis, message=FALSE, warning=FALSE}
    # merge the chronic diseases data with the county type data
    chronic_diseases_data <- merge(chronic_disease_data, enriched_county_data, by = 'LocationID')

    # we will rename urban_rural_classification to urban_rural_sub_class
    # and based on the values of urban_rural_code, we will create a new column urban_rural_classification as follows:
    # 1 -> Urban, 2 -> Urban, 3 -> Urban, 4 -> Urban, 5 -> Rural, 6 -> Rural
    chronic_diseases_data$urban_rural_sub_class <- chronic_diseases_data$urban_rural_classification
    chronic_diseases_data$urban_rural_classification <- ifelse(chronic_diseases_data$urban_rural_code %in% c(1, 2, 3, 4), 'Urban', 'Rural')

    # calculate the mean age-adjusted prevalence of chronic diseases (chronic_disease_cols) by urban_rural_classification
    chronic_diseases_mean <- chronic_diseases_data %>% group_by(urban_rural_classification) %>% summarise(across(all_of(chronic_disease_cols), mean, na.rm = TRUE))

    # display the mean age-adjusted prevalence of chronic diseases by urban_rural_classification
    chronic_diseases_mean

    # visualize the mean age-adjusted prevalence of chronic diseases by urban_rural_classification
    # box plot for each chronic disease column by urban_rural_classification
    library(ggplot2)
    library(tidyr)

    chronic_diseases_data_long <- chronic_diseases_data %>% pivot_longer(cols = chronic_disease_cols, names_to = 'ChronicDisease', values_to = 'AgeAdjustedPrevalence')

    ggplot(chronic_diseases_data_long, aes(x = urban_rural_classification, y = AgeAdjustedPrevalence, fill = urban_rural_classification)) +
        geom_boxplot() +
        facet_wrap(~ChronicDisease, scales = 'free_y') +
        labs(title = 'Mean Age-Adjusted Prevalence of Chronic Diseases by County Type', x = 'County Type', y = 'Age-Adjusted Prevalence') +
        theme_minimal()

    # perform a t-test to determine if there is a significant difference in the age-adjusted prevalence of chronic diseases between urban and rural counties
    t_test_results <- lapply(chronic_disease_cols, function(col) {
        t_test <- t.test(chronic_diseases_data[[col]] ~ chronic_diseases_data$urban_rural_classification)
        data.frame(ChronicDisease = col, p_value = t_test$p.value, mean_urban = t_test$estimate[1], mean_rural = t_test$estimate[2])
    }) %>% bind_rows()

    # display the t-test results
    t_test_results
```

