---
title: "Chronic Disease Mortality - Predictive Modeling"
author: "Chetan, Augustine"
output: 
  html_document:
    keep_md: true
    df_print: paged
---
    
    ```{r setup, include=FALSE}
       knitr::opts_chunk$set(echo = TRUE)
    ```

# Introduction
Using CDC's Chronic Disease Indicators dataset, we are interested in analyzing 
the relationship between _chronic disease mortality rate_ and _prevalence of preventive care_.

These are datasets we are using for our analysis:
1. CDC Chronic Disease Indicators (CDI) - https://www.cdc.gov/cdi/
2. PLACES: Local Data for Better Health - https://www.cdc.gov/places/index.html

# Research Question
**Does _low prevalence of preventive care_ lead to _higher chronic disease mortality rate_?**

# Hypothesis
We hypothesize that low prevalence of preventive care leads to higher chronic disease mortality rate.

# Methodology
We will use predictive modeling to analyze the relationship between chronic disease mortality rate and prevalence of preventive care. 

We will use the following steps to conduct our analysis:
1. Data Preprocessing
2. Exploratory Data Analysis
3. Model Development
4. Model Evaluation
5. Conclusion

# Data
```{r data, echo=TRUE}
library(readr)
library(dplyr)

# # Load the data for mortality rate from chronic diseases
# cdi_dataset <- read.csv("us_cdi_dataset.csv", stringsAsFactors = FALSE)

# # above dataset is downloaded from here: https://data.cdc.gov/Chronic-Disease-Indicators/U-S-Chronic-Disease-Indicators/hksd-2xuw/about_data
# # it's about 400MB in size, so I am not uploading it to the repository

# # From this dataset, we are only interested in extracting mortality rate for chronic diseases

# mortality_data <- cdi_dataset %>% 
#   filter(grepl("mortality", Question, ignore.case = TRUE)) %>%
#     filter(DataValueType == "Age-adjusted Rate") %>%
#         select(Topic, Question, DataValueType, DataValueUnit, StratificationCategory1, Stratification1, 
#           DataValue, DataSource, LocationDesc, LocationAbbr, GeoLocation, LocationID, YearStart, YearEnd)

# # number of rows and columns
# dim(mortality_data)

# head(mortality_data)

# # lets write this data to a csv file (we will use this data for our analysis)
# write.csv(mortality_data, "mortality_data.csv", row.names = TRUE)

# # Load the data for prevalence of preventive care
# places_dataset <- read.csv("PLACES__Local_Data_for_Better_Health__Place_Data_2023_release_20240314.csv", stringsAsFactors = FALSE)

# # above dataset is downloaded from here: https://data.cdc.gov/500-Cities-Places/PLACES-Local-Data-for-Better-Health-Place-Data-202/eav7-hnsx/about_data
# # it's about 500MB in size, so I am not uploading it to the repository

# # From this dataset, we are only interested in extracting prevalence of preventive care
# # Category = Prevention and Data_Value_Type = Age-adjusted prevalence

# preventive_care_data <- places_dataset %>% 
#   filter(Category == "Prevention") %>%
#     filter(Data_Value_Type == "Age-adjusted prevalence") %>%
#         select(Category, Short_Question_Text, Measure, Data_Value_Type, 
#             Data_Value_Unit, Data_Value, StateDesc, StateAbbr, Geolocation, LocationID, Year)

# # number of rows and columns
# dim(preventive_care_data)

# head(preventive_care_data)

# # lets write this data to a csv file (we will use this data for our analysis)
# write.csv(preventive_care_data, "preventive_care_data.csv", row.names = TRUE)
```

# Data Preprocessing
```{r data_preprocessing, echo=TRUE}
# mortality data
# Load the data (row names are included in the dataset)
mortality_data <- read.csv("mortality_data.csv", stringsAsFactors = FALSE, row.names = 1)
# mortality_data <- as.data.frame(mortality_data)
head(mortality_data)

# Number of rows and columns
dim(mortality_data)

# Check for missing values
missing_values <- mortality_data %>% 
  summarise_all(funs(sum(is.na(.))))
missing_values

# What's the percentage of missing values?
missing_values_percentage <- missing_values / nrow(mortality_data) * 100
missing_values_percentage

# preventive care data
# Load the data (row names are included in the dataset)
preventive_care_data <- read.csv("preventive_care_data.csv", stringsAsFactors = FALSE, row.names = 1)
# preventive_care_data <- as.data.frame(preventive_care_data)
head(preventive_care_data)

# Number of rows and columns
dim(preventive_care_data)

# Check for missing values
missing_values <- preventive_care_data %>% 
  summarise_all(funs(sum(is.na(.))))
missing_values

# What's the percentage of missing values?
missing_values_percentage <- missing_values / nrow(preventive_care_data) * 100
missing_values_percentage

# unique values for Topic and Question in mortality data
unique(mortality_data$Topic)
# unique(mortality_data$Question)

# unique values for Short_Question_Text and Measure in preventive care data
unique(preventive_care_data$Short_Question_Text)
# unique(preventive_care_data$Measure)
```
