---
title: "Chronic Disease Mortality - Predictive Modeling"
author: "Chetan, Augustine"
output: 
  html_document:
    keep_md: true
    df_print: paged
---
    
    ```{r setup, include=FALSE}
       knitr::opts_chunk$set(echo = TRUE)
    ```

# Introduction
Using CDC's Chronic Disease Indicators dataset, we are interested in analyzing if higher prevalence of preventive care leads to lower chronic disease mortality rate. 
We will use predictive modeling to analyze the relationship between chronic disease mortality rate and prevalence of preventive care.

These are datasets we are using for our analysis:
1. CDC Chronic Disease Indicators (CDI) - https://www.cdc.gov/cdi/
2. PLACES: Local Data for Better Health - https://www.cdc.gov/places/index.html

# Research Question
**What is the relationship between chronic disease mortality rate and prevalence of preventive care?**

# Hypothesis
We hypothesize that low prevalence of preventive care leads to higher chronic disease mortality rate.

Null Hypothesis: There is no relationship between chronic disease mortality rate and prevalence of preventive care.

# Methodology
We will use predictive modeling to analyze the relationship between chronic disease mortality rate and prevalence of preventive care. We will use the following steps:
1. Mortality Rate Data: Extract mortality rate data from CDC Chronic Disease Indicators dataset.
2. Prevalence of Preventive Care Data: Extract prevalence of preventive care data from PLACES dataset.
3. Data Preprocessing: Check for missing values, data types, and unique values in the datasets.
4. Data Analysis: Analyze the relationship between chronic disease mortality rate and prevalence of preventive care.
5. Predictive Modeling: Use machine learning algorithms to predict chronic disease mortality rate based on prevalence of preventive care.
6. Model Evaluation: Evaluate the performance of the predictive models using metrics like RMSE, MAE, and R-squared.
7. Conclusion: Draw conclusions based on the predictive modeling results.

# Data
```{r data, echo=TRUE}
library(readr)
library(dplyr)

# # Load the data for mortality rate from chronic diseases
# cdi_dataset <- read.csv("us_cdi_dataset.csv", stringsAsFactors = FALSE)

# # above dataset is downloaded from here: https://data.cdc.gov/Chronic-Disease-Indicators/U-S-Chronic-Disease-Indicators/hksd-2xuw/about_data
# # it's about 400MB in size, so I am not uploading it to the repository

# # From this dataset, we are only interested in extracting mortality rate for chronic diseases

# mortality_data <- cdi_dataset %>% 
#   filter(grepl("mortality", Question, ignore.case = TRUE)) %>%
#     filter(DataValueType == "Age-adjusted Rate") %>%
#         select(Topic, Question, DataValueType, DataValueUnit, StratificationCategory1, Stratification1, 
#           DataValue, DataSource, LocationDesc, LocationAbbr, GeoLocation, LocationID, YearStart, YearEnd)

# # number of rows and columns
# dim(mortality_data)

# head(mortality_data)

# # lets write this data to a csv file (we will use this data for our analysis)
# write.csv(mortality_data, "mortality_data.csv", row.names = TRUE)

# # Load the data for prevalence of preventive care
# places_dataset <- read.csv("PLACES__Local_Data_for_Better_Health__Place_Data_2023_release_20240314.csv", stringsAsFactors = FALSE)

# # above dataset is downloaded from here: https://data.cdc.gov/500-Cities-Places/PLACES-Local-Data-for-Better-Health-Place-Data-202/eav7-hnsx/about_data
# # it's about 500MB in size, so I am not uploading it to the repository

# # From this dataset, we are only interested in extracting prevalence of preventive care
# # Category = Prevention and Data_Value_Type = Age-adjusted prevalence

# preventive_care_data <- places_dataset %>% 
#   filter(Category == "Prevention") %>%
#     filter(Data_Value_Type == "Age-adjusted prevalence") %>%
#         select(Category, Short_Question_Text, Measure, Data_Value_Type, 
#             Data_Value_Unit, Data_Value, StateDesc, StateAbbr, Geolocation, LocationID, Year)

# # number of rows and columns
# dim(preventive_care_data)

# head(preventive_care_data)

# # lets write this data to a csv file (we will use this data for our analysis)
# write.csv(preventive_care_data, "preventive_care_data.csv", row.names = TRUE)
```

# Data Preprocessing
```{r data_preprocessing, echo=TRUE}
# mortality data
# Load the data (row names are included in the dataset)
mortality_data <- read.csv("mortality_data.csv", stringsAsFactors = FALSE, row.names = 1)
# mortality_data <- as.data.frame(mortality_data)
head(mortality_data)

# Number of rows and columns
dim(mortality_data)

# Check for missing values
missing_values <- mortality_data %>% 
  summarise_all(funs(sum(is.na(.))))
missing_values

# What's the percentage of missing values?
missing_values_percentage <- missing_values / nrow(mortality_data) * 100
missing_values_percentage

# preventive care data
# Load the data (row names are included in the dataset)
preventive_care_data <- read.csv("preventive_care_data.csv", stringsAsFactors = FALSE, row.names = 1)
# preventive_care_data <- as.data.frame(preventive_care_data)
head(preventive_care_data)

# Number of rows and columns
dim(preventive_care_data)

# Check for missing values
missing_values <- preventive_care_data %>% 
  summarise_all(funs(sum(is.na(.))))
missing_values

# What's the percentage of missing values?
missing_values_percentage <- missing_values / nrow(preventive_care_data) * 100
missing_values_percentage

# unique values for Topic and Question in mortality data
unique(mortality_data$Topic)
# unique(mortality_data$Question)

# unique values for Short_Question_Text and Measure in preventive care data
unique(preventive_care_data$Short_Question_Text)
# unique(preventive_care_data$Measure)
```
